Complete draft

### 2019-08-31 doc/point.tex
snapshot

Suppose we have two sets of data, the number of galaxies $N$ in a volume
bin and the radial velocities of galaxies $v$.  These may or may not be the same galaxies.
The likelihood is then
\begin{align}
L(\theta; N, v) & =p(N, v| \theta) \\
&= \int  p(N | \delta)  p(v | u)  p(\delta,v| \theta) d\delta\, du\\ 
& = \int  \prod_{i=1}^{N_{bin}}   p(N_i | \delta_i)p(v| u) p(\delta, u | \theta)d\delta\, du\\ 
& = \int  \prod_{i=1}^{N_{bin}}  \sum_{\tilde{N}_i =N_i}^\infty  p(N_i | \tilde{N}_i) p(\tilde{N}_i | \delta) p(v| u) p(\delta, u | \theta) d\delta\, du \\
& = \int  \prod_{i=1}^{N_{bin}}  \sum_{\tilde{N}_i =N_i}^\infty  \mathcal{B}(N_i; \tilde{N}_i, p_i) \mathcal{P}(\tilde{N}_i ; N_{0i}(\delta))p(v| u) p(\delta, u | \theta) d\delta\, du \\
&= \int  \exp \left[ \sum_{i=1}^{N_{bin}} \ln{  \left( \sum_{\tilde{N}_i =N_i}^\infty  \mathcal{B}(N_i; \tilde{N}_i, p_i) \mathcal{P}(\tilde{N}_i ; N_{0i}(\delta))\right)} \right]  p(v| u) p(\delta, u | \theta) d\delta\, du
\end{align}
Here $\delta$ and $i$ are  the mass-density  and momentum fields at the bin/galaxy positions, $\tilde{N}_i$ is the latent parameter
for the  total number of galaxies in bin $i$, $N_{0i}(\hat{\delta})$ is the expected
average number of galaxies in bin $i$, and $p_i$ is the probability of for a galaxy in cell $i$ to enter the sample.
$\mathcal{B}$ is the binomial distribution and $\mathcal{P}$ is the Poisson distribution.

The summation over bins an be split up into two terms assuming there are no cells with greater than two galaxies.
The first is a term as if there were zero galaxies in all bins, $N_i=0\ \forall i$.
\begin{align}
 \sum_{i=1}^{N_{bin}} \ln{  \left( \sum_{\tilde{N}_i =N_i}^\infty  \mathcal{B}(0; \tilde{N}_i, p_i) \mathcal{P}(\tilde{N}_i ; N_{0i}(\delta))\right)} & =
\sum_{i=1}^{N_{bin}} \ln{  \left( \sum_{\tilde{N}_i =0}^\infty   (1-p_i)^{ \tilde{N}_i} \frac{N_{0i}^{\tilde{N}_i} e^{-N_{0i}}}{\tilde{N}_i!} \right)}\\ 
& = \sum_{i=1}^{N_{bin}} (1-p_i)N_{0i} - N_{0i}\\
& = - \sum_{i=1}^{N_{bin}} p_i N_{0i}  \\
& = - \int  p(x) n_{0}(x) d^3x
\end{align}
where in the limit of  infinitesimally small bins $N_{0}$ approaches zero and
the integral replaces the summation.
The second term is over those cells that have one galaxy:
\begin{align}
\sum_{i \in \{N_i \ne 0\}}
\ln{  \left( \sum_{\tilde{N}_i =N_i}^\infty  \mathcal{B}(N_i; \tilde{N}_i, p_i) \mathcal{P}(\tilde{N}_i ; N_{0i}(\delta))\right)} -
 \ln{  \left( \sum_{\tilde{N}_i =N_i}^\infty  \mathcal{B}(0; \tilde{N}_i, p_i) \mathcal{P}(\tilde{N}_i ; N_{0i}(\delta))\right)}\\
 =\sum_{i \in \{N_i \ne 0\}} \ln{  \left( \sum_{\tilde{N}_i =1}^\infty   \tilde{N}_i p_i (1-p_i)^{ \tilde{N}_i-1} \frac{N_{0i}^{\tilde{N}_i} e^{-N_{0i}}}{\tilde{N}_i!} \right)}
- \ln{  \left( \sum_{\tilde{N}_i =0}^\infty   (1-p_i)^{ \tilde{N}_i} \frac{N_{0i}^{\tilde{N}_i} e^{-N_{0i}}}{\tilde{N}_i!} \right)} \\
 = \sum_{i=1}^{N_{\text{galaxies}}} \ln{  \left(  p_i  {N_{0i} } \right)} \\
  = \sum_{i=1}^{N_{\text{galaxies}}} \ln{  \left(  p_i  n_{0}(x_i) \hat{\delta} V \right)}
\end{align}

Then the likelihood is
\begin{align}
L(\theta; N, v) 
&\propto \int  \exp \left(- \int  p(x) n_{0}(x) d^3x\right)  \prod_{i=1}^{N_{\text{galaxies}}} p_i n_0(x_i) p(v_i| u_i) p(\delta, u | \theta) d\delta\, du .
\end{align}
The exponential term depends on $\delta$ at every differential volume.  The
remaining terms of the integrand depend only on the $\delta_i$ and $u_i$ at galaxy coordinates.

When the universe is modeled as a periodic cube much larger than the survey. the the density field can be describe by discrete points. 
Further assuming a cutoff in power at $k_{max}$, the field can be described by a finite grid.  Then the parameters $\delta$ and $u$ will be
the Fourier density and momentum at grid points, having a covariance matrix composed of four diagonal submatrices. 
The density and momentum at any point in real space can be constructed from the parameters.

>>>>

Suppose we have two sets of data, the number of galaxies $N$ in a volume
bin and the radial velocities of galaxies $v$.  These may or may not be the same galaxies.
We also assume that the universe is periodic with a minimum length scale, meaning that fields can be
described by a finite number of Fourier modes.  Given a model of gravity, we can deterministically calculate the
velocity field given the density field.
The likelihood is then
\begin{align}
L(\theta; N, v) & =p(N, v| \theta) \\
& = \int p(N,v,\delta | \theta) d\delta \\\
& = p(v|\theta)  \int p(N | \delta) p(\delta| v, \theta) d\delta \\
& = p(v|\theta)  \int \prod_{i=1}^{N_{bin}}  p(N_i | \delta) p(\delta| v, \theta) d\delta \\
& = p(v|\theta)  \int \prod_{i=1}^{N_{bin}}  \sum_{\tilde{N}_i =N_i}^\infty    p(N_i | \tilde{N}_i) p(\tilde{N}_i | \delta)  p(\delta| v, \theta) d\delta \\
& = p(v|\theta)  \int \prod_{i=1}^{N_{bin}}  \sum_{\tilde{N}_i =N_i}^\infty    \mathcal{B}(N_i; \tilde{N}_i, p_i) \mathcal{P}(\tilde{N}_i ; N_{0i}(\delta))  p(\delta| v, \theta) d\delta \\
& =  p(v|\theta)  \int  \exp \left[ \sum_{i=1}^{N_{bin}} \ln{  \left( \sum_{\tilde{N}_i =N_i}^\infty  \mathcal{B}(N_i; \tilde{N}_i, p_i) \mathcal{P}(\tilde{N}_i ; N_{0i}(\delta))\right)} \right]  p(\delta| v, \theta) d\delta 
%&= \int  p(N | \delta)  p(v | u)  p(\delta,v| \theta) d\delta\, du\\ 
%& = \int  \prod_{i=1}^{N_{bin}}   p(N_i | \delta_i)p(v| u) p(\delta, u | \theta)d\delta\, du\\ 
%& = \int  \prod_{i=1}^{N_{bin}}  \sum_{\tilde{N}_i =N_i}^\infty  p(N_i | \tilde{N}_i) p(\tilde{N}_i | \delta) p(v| u) p(\delta, u | \theta) d\delta\, du \\
%& = \int  \prod_{i=1}^{N_{bin}}  \sum_{\tilde{N}_i =N_i}^\infty  \mathcal{B}(N_i; \tilde{N}_i, p_i) \mathcal{P}(\tilde{N}_i ; N_{0i}(\delta))p(v| u) p(\delta, u | \theta) d\delta\, du \\
%&= \int  \exp \left[ \sum_{i=1}^{N_{bin}} \ln{  \left( \sum_{\tilde{N}_i =N_i}^\infty  \mathcal{B}(N_i; \tilde{N}_i, p_i) \mathcal{P}(\tilde{N}_i ; N_{0i}(\delta))\right)} \right]  p(v| u) p(\delta, u | \theta) d\delta\, du
\end{align}
Here $\delta$ and $i$ are  the mass-density  and momentum fields at the bin/galaxy positions, $\tilde{N}_i$ is the latent parameter
for the  total number of galaxies in bin $i$, $N_{0i}(\hat{\delta})$ is the expected
average number of galaxies in bin $i$, and $p_i$ is the probability of for a galaxy in cell $i$ to enter the sample.
$\mathcal{B}$ is the binomial distribution and $\mathcal{P}$ is the Poisson distribution.

The summation over bins an be split up into two terms assuming there are no cells with greater than two galaxies.
The first is a term as if there were zero galaxies in all bins, $N_i=0\ \forall i$.
\begin{align}
 \sum_{i=1}^{N_{bin}} \ln{  \left( \sum_{\tilde{N}_i =N_i}^\infty  \mathcal{B}(0; \tilde{N}_i, p_i) \mathcal{P}(\tilde{N}_i ; N_{0i}(\delta))\right)} & =
\sum_{i=1}^{N_{bin}} \ln{  \left( \sum_{\tilde{N}_i =0}^\infty   (1-p_i)^{ \tilde{N}_i} \frac{N_{0i}^{\tilde{N}_i} e^{-N_{0i}}}{\tilde{N}_i!} \right)}\\ 
& = \sum_{i=1}^{N_{bin}} (1-p_i)N_{0i} - N_{0i}\\
& = - \sum_{i=1}^{N_{bin}} p_i N_{0i}  \\
& = - \int  p(x) n_{0}(x) d^3x
\end{align}
where in the limit of  infinitesimally small bins $N_{0}$ approaches zero and
the integral replaces the summation.

Consider a cell with $N_i$ galaxies.  It contributes a 
\begin{align}
\ln{  \left( \sum_{\tilde{N}_i =N_i}^\infty  \mathcal{B}(N_i; \tilde{N}_i, p_i) \mathcal{P}(\tilde{N}_i ; N_{0i}(\delta))\right)}\\
= \ln{  \left( \sum_{\tilde{N}_i =N_i}^\infty \frac{\tilde{N}_i!}{N_i!(\tilde{N}_i-N_i)!}   p_i^{N_i} (1-p_i)^{ \tilde{N}_i-N_i} \frac{N_{0i}^{\tilde{N}_i} e^{-N_{0i}}}{\tilde{N}_i!} \right)}
 \\
 = \ln{  \left( \frac{(p_iN_{0i})^N_ie^{-p_iN_{0i}}}{N_i!}  \right)} \\
 =  N_i \ln{\left(  p_i  N_{0i} \right) } -p_i N_{0i} - \ln{N_i!}.
\end{align}
For those cells that have a galaxy subtract out the probability for zero objects.  Each of these cells thus contribute
\begin{equation}
 N_i \ln{\left(  p_i  N_{0i} \right) } - \ln{N_i!}.
\end{equation}

The exponent term is proportional to (we don't care about multiplicative terms independent of the parameters)
\begin{align}
\exp\left(- \sum_{i=1}^{N_{bin}} p_i N_{0i}  \right)
 \prod_{i=1}^{N_{bin}}  (p_i N_{0i})^{N_i} 
\end{align}

The second term is over those cells that have one galaxy:
\begin{align}
\sum_{i \in \{N_i \ne 0\}}
\ln{  \left( \sum_{\tilde{N}_i =N_i}^\infty  \mathcal{B}(N_i; \tilde{N}_i, p_i) \mathcal{P}(\tilde{N}_i ; N_{0i}(\delta))\right)} -
 \ln{  \left( \sum_{\tilde{N}_i =N_i}^\infty  \mathcal{B}(0; \tilde{N}_i, p_i) \mathcal{P}(\tilde{N}_i ; N_{0i}(\delta))\right)}\\
 =\sum_{i \in \{N_i \ne 0\}} \ln{  \left( \sum_{\tilde{N}_i =1}^\infty   \tilde{N}_i p_i (1-p_i)^{ \tilde{N}_i-1} \frac{N_{0i}^{\tilde{N}_i} e^{-N_{0i}}}{\tilde{N}_i!} \right)}
- \ln{  \left( \sum_{\tilde{N}_i =0}^\infty   (1-p_i)^{ \tilde{N}_i} \frac{N_{0i}^{\tilde{N}_i} e^{-N_{0i}}}{\tilde{N}_i!} \right)} \\
 = \sum_{i=1}^{N_{\text{galaxies}}} \ln{  \left(  p_i  {N_{0i} } \right)} \\
  = \sum_{i=1}^{N_{\text{galaxies}}} \ln{  \left(  p_i  n_{0}(x_i) \delta V \right)}
\end{align}

Then the likelihood is
\begin{align}
L(\theta; N, v) 
&\propto p(v|\theta)  \int \exp \left(- \int  p(x) n_{0}(x) d^3x\right)  \prod_{i=1}^{N_{\text{galaxies}}} p(x_i) n_0(x_i)  p(\delta| v, \theta) d\delta \\
& \approx p(v|\theta)  \int \prod_{j=1}^{N_\text{grid}} \exp \left( - \frac{p(x_j) n_{0}(x_j)}{N_\text{grid}}\right) \prod_{i=1}^{N_{\text{galaxies}}} p(x_i) n_0(x_i)  p(\delta| v, \theta) d\delta 
%&\propto \int  \exp \left(- \int  p(x) n_{0}(x) d^3x\right)  \prod_{i=1}^{N_{\text{galaxies}}} p(x_i) n_0(x_i) p(v_i| u_i) p(\delta, u | \theta) d\delta\, du \\
%& \approx  \int  \prod_{j=1}^{N_\text{grid}} \exp \left( - \frac{p(x_j) n_{0}(x_j)}{N_\text{grid}}\right) \prod_{i=1}^{N_{\text{galaxies}}} p(x_i) n_0(x_i) p(v_i| u_i) p(\delta, u | \theta) d\delta\, du.
\end{align}
The exponential term depends on $\delta$ at every differential volume.  The
remaining terms of the integrand depend only on the $\delta_i$ and $u_i$ at galaxy coordinates.

When the universe is modeled as periodic, the fields therein can be expanded as a Fourier series.  Truncating the expansion at some $k_{max}$
means that the a field can be described by a finite set of numbers.  The parameters $\delta$ and $u$ will be
the Fourier density and momentum at grid points.  Realizations of these parameters can be easily generated because their 
 covariance matrix is composed of four diagonal submatrices. 
The density and momentum at any point in real space can be constructed from the parameters using the Fourier transform.
